---
title: "Integration of RNAseq Analysis tools with RNAseqEasy"
author: Salvador Torres Montilla
date: "Last edited `r format(Sys.time(), '%d %B, %Y')`"
abstract: >
  RNAseq analysis has become a common practice in most molecular biology
  laboratories in the world. Several tools have been developed to achieve 
  differential expression analysis, functional enrichment, gene co-expression 
  or data visualization among others, from RNAseq data. However, many of them
  require independent analysis and, in many ocasions, relatively high
  programming skills. The package RNAseqEasy provides an integration of some of
  this tools into a single user-friendly workflow, and tries to simplify it to 
  require only basic notions in R programming. It is mainly based on DESeq2 
  [@DESeq2], topGO [@topGO] and WGCNA [@WGCNA] packages to carry out PCA
  exploratory plots, differential expression analysis, gene ontology enrichment 
  analysis and gene-coexpression from Salmon [@patro2017] quantified data. This 
  vignette explains the use of the package and demonstrates typical workflows.
  
  
  RNAseqEasy package version: `r packageVersion("RNAseqEasy")`
output: 
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Integration of RNAseq Analysis tools with RNAseqEasy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
bibliography: references.bib
csl: nature.csl
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
temp_output_dir <- tempdir()
library(RNAseqEasy)
```

## Input data

The RNAseqEasy package is thought to be used from Salmon mapped and quantified data. The first required input is the path of the directory where `quant.sf` samples are saved will be the input, and the different functions included in this package will retrieve the data from it. The other input necessary for the analysis is a data frame that includes the name of the samples and the information describing them (i.e. variables or conditions defining the experiment).

### Salmon data

Salmon [@patro2017] is a widely used software for quantifying transcript abundance. A well detailed tutorial can be found in their [official webpage](https://combine-lab.github.io/salmon/getting_started/). It is a very fast tool for transcript quantification directly from `fastq.gz`files. The only requirement is the creation of an index of the **transcriptome** of the organism we are working with (do not use the genome!!!), also explained in their tutorial.

As a result, we will obtain a folder [1] containing one folder per analyzed sample with its names. In each folder, the main output fail will be named `quant.sf`, which contains the name of each transcript and their abundance in Transcripts Per Million (TPM), among other information (length, effective length and number of reads). The path to this output folder [1] will be input for our analysis.



```{r zenodo}
# URL from Zenodo
data_url <- "https://zenodo.org/records/15800134/files/GSE275561.zip?download=1"

# Temporal directory to download data
output_dir <- file.path(tempdir(), "GSE275561")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

zip_file <- file.path(tempdir(), "GSE275561.zip")

# Download zip file with Salmon quantified data from Zenodo
download.file(url = data_url, destfile = zip_file, mode = "wb")

# Unzip compressed file
unzip(zip_file, exdir = output_dir)

list.files(output_dir)
```

