---
title: "RNAseqEasy: GO Enrichment Analysis with topGO"
author: Salvador Torres Montilla
date: "Last edited `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{RNAseqEasy: GO Enrichment Analysis with topGO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
bibliography: references.bib
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
temp_output_dir <- tempdir()
```

## Introduction

This vignette demonstrates how to use the `RNAseqEasy` package to perform GO enrichment analysis using the `topGO_All()` function. We will use a toy dataset to illustrate the workflow. The main goal is to identify GO terms representing a list of Genes (e.g., differential expressed genes) and representing the output in useful plots.

The workflow if focused on the `topGO_All()` function, which integrates different steps: preparing the gene data, executing the enrichment analysis using `topGO` package, processing the results and visualizing the data.

First, we import the require packages. `RNAseqEasy` is our package, and 'topGO' a key dependency for the analysis.

```{r load_packages}
library(RNAseqEasy)
library(dplyr)
```

## Data preparation

To perform GO enrichment analysis, we need two principal components:

1.  **A list of genes of interest**: It may be a Gene IDs vector or a `data.frame` resulting from a differential gene enrichment analysis (which may include Gene IDs as rownames and, optionally, a `padj` column including significance data for filtering)-
2.  **Functional annotation of GO terms (`geneID2GO`)**: A named list where each element corresponds to a gene and contains a vector with GOterms associated to this gene.

`RNAseqEasy` package includes example data to illustrate this process:

-   `DEG_example_list`: A `data.frame` resulting from `DESeq2` analysis, comparing *gh3a* mutant vs Tak-1 in Mock conditions in *Marchantia polymorpha* [@liang2024].
-   `Mpo_GO_GOSLIM`: A `data.frame` containing GO annotations for *Marchantia polymorpha* transcripts. Relevant columns are "Transcript" (Gene IDs) and "GO" (associated GO term).

We load this example data:

```{r load_example_data}
data("DEG_example_list", package = "RNAseqEasy")
data("Mpo_GO_GOSLIM", package = "RNAseqEasy")

head(DEG_example_list)
head(Mpo_GO_GOSLIM)
```

The function *load_topGO_db* can be used to convert `Mpo_GO_GOSLIM` object into a `geneID2GO` list required for *topGO* analysis.

```{r create_geneID2GO}
geneID2GO <- load_topGO_db(Mpo_GO_GOSLIM, GeneID =  "Transcript", GOterms = "GO")

# Verificamos la estructura de geneID2GO (debería ser una lista nombrada)
str(head(geneID2GO))
```

